---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First we extract the activity data from the zipped archive.
```{r loadingdata}
temp<-unz("activity.zip","activity.csv")
mydata<-read.csv(temp,header=TRUE)
unlink(temp)
remove(temp)
```

## What is mean total number of steps taken per day?
```{r meansteps, echo=FALSE, results="asis"}
temp <- mydata[!is.na(mydata$steps),]
stepsByDate <- data.frame(with(temp, tapply(steps,date,sum,na.rm=TRUE)))
names(stepsByDate) <- c("steps")

options(digits=2)
steps <- as.integer(mean(stepsByDate$steps,na.rm = TRUE))
stepsmedian <- as.integer(median(stepsByDate$steps,na.rm=TRUE))
```

The **mean** number of steps is `r steps`, and the **median** is `r stepsmedian`. 

```{r histogram, echo=FALSE}
hist(stepsByDate$steps,main="Histogram of Total Daily Steps", xlab="Total Steps")
abline(v=steps)
```

## What is the average daily activity pattern?

```{r dailyactivity, echo=FALSE, message=FALSE}
library(dplyr)
x <- group_by(mydata,interval)
y <- x %>% summarize(mean=mean(steps,na.rm=TRUE))
maxsteps <- as.integer(y[which(y$mean==max(y$mean)),"interval"])
```

The **maximum number of steps** occur in interval `r maxsteps`.

```{r plotforsteps, echo=FALSE}
with(y,plot(interval,mean,type="l",main="Average Daily Pattern",xlab="Interval",ylab="Mean Steps"))
```



## Imputing missing values
```{r impute, echo=FALSE, message=FALSE}
numNAs <- sum(is.na(mydata$steps))

library(plyr)   #for ddply
library(Hmisc)  #for impute
mydata2<-ddply(mydata,"interval",mutate,imputed.value=impute(steps,mean))

stepsByDate <- data.frame(with(mydata2, tapply(imputed.value,date,sum,na.rm=TRUE)))
names(stepsByDate) <- c("steps")
options(digits=2)
steps <- as.integer(mean(stepsByDate$steps,na.rm = TRUE))
stepsmedian <- as.integer(median(stepsByDate$steps,na.rm=TRUE))

```

There are `r numNAs` missing values in the dataset. 

In the imputed dataset, the **mean** number of steps is `r steps`, and the **median** is `r stepsmedian`. These values are nearly identical to the first part of the assignments. The imputations did not alter the mean/median values (though appears to have brought them closer together).

```{r histogramimputed, echo=FALSE}
hist(stepsByDate$steps,main="Histogram of Total Daily Steps in IMPUTED Dataset", xlab="Total Steps")
abline(v=steps)
```

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=FALSE, message=FALSE, warning=FALSE}
mydata2 <- mutate(mydata2,weekday=weekdays(as.Date(date)))
weekdata <- subset(mydata2,weekday %in% c("Monday","Wednesday","Tuesday","Thursday","Friday"))
weekenddata <- subset(mydata2,weekday %in% c("Saturday","Sunday"))

names(weekdata)<-c("steps","date","interval","steps2","weekday")
names(weekenddata)<-c("steps","date","interval","steps2","weekday")

library(dplyr)
detach("package:plyr", unload=TRUE)
y1 <- weekdata %>% group_by(interval) %>% summarise(mean=mean(steps2))
y2 <- weekenddata %>% group_by(interval) %>% summarise(mean=mean(steps2))
finaldata <- rbind(mutate(y1,week="weekday"),mutate(y2,week="weekend"))
```

```{r plotforsteps2, echo=FALSE}
xyplot(mean~interval|week,data=finaldata,type="l",layout=c(1,2))
```
